Elucidating the Transcriptome of Turkey Hemorrhagic Enteritis Virus

### 2023-12-20 manuscript_thev_transcriptome.Rmd
made fig 3 as one plot and removed uneeded plots

---
title: Elucidating the Transcriptome of Turkey Hemorrhagic Enteritis Virus
output:
  word_document: default
  pdf_document: default
geometry: margin=1in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
bibliography: transcriptome_refs.bib
csl: asm.csl
editor_options:
  chunck_output_type: console
---
\vspace{5mm}

@@

\vspace{10mm}
Abraham Quaye^1\*^, Bret Pickett^\*^, Joel S. Griffitts^\*^, Bradford K. Berges^\*^, Brian D. Poole${^\dagger}$^\*^

@@

^\*^Department of Microbiology and Molecular Biology, Brigham Young University       
^1^First-author      
${^\dagger}$ Corresponding Author

\vspace{5mm}
**Corresponding Author Information**     
[brian_poole@byu.edu](mailto:brian_poole@byu.edu)       
Department of Microbiology and Molecular Biology,       
4007 Life Sciences Building (LSB),    
Brigham Young University,   
Provo, Utah  
\newpage

@@

most_sig_72 <- bulk_junc_stats %>% filter(timepoint == "72hpi", region == "MLP") %>% pull(read_count) %>% sum()/as.numeric(tab1$`72h.p.i`[6])*100
```
## ABSTRACT        
**Background:** Hemorrhagic enteritis (HE) is a disease affecting 6-12-week-old turkeys characterized by _immunosuppression (IS)_ and bloody diarrhea. This disease is caused by _Turkey Hemorrhagic Enteritis Virus (THEV)_ of which avirulent strains (THEV-A) that do not cause HE but retain the immunosuppressive ability have been isolated. The THEV-A Virginia Avirulent Strain (VAS) is still used as a live vaccine despite its immunosuppressive properties. _Our objective is to understand the genetic basis by which VAS induces IS_. The _transcriptome of THEV_ was studied to set the stage for further experimentation with specific viral genes that may mediate IS.      
**Methods:** After infecting a turkey B-cell line (MDTC-RP19) with the VAS vaccine strain, samples in triplicates were collected at 4-, 12-, 24-, and 72-hours post-infection. Total RNA was subsequently extracted, and poly-A-tailed mRNA sequencing done. After trimming the raw sequencing reads with the `FastQC`, reads were mapped to the THEV genome using `Hisat2` and transcripts assembled with `StringTie`. An in-house script was used to consolidate transcripts from all time-points, generating the final transcriptome. PCR, gel electrophoresis, and Sanger sequencing were used to validate all identified splice junctions.      
**Results and Conclusions:** A total of `r total_mapped` million reads mapped to THEV genome providing good coverage/depth, leaving no regions unmapped. All predicted genes in the genome were represented. In keeping with all adenoviruses, all transcripts were spliced with either with 5'- or 3'-multi exon UTRs hitherto unknown. *~~Thirteen~~* novel exons were identified which were validated by PCR and Sanger sequencing. The splicing patterns strongly suggest that there are *~~three~~* main promoters (E1, E3, and major late promoters) driving expression of most of the genes with *~~two~~* possible minor promoters driving single genes (ORF7 and ORF8). This RNA-sequencing experiment is the first study of THEV gene expression to date. In keeping with other Adenoviruses, almost all THEV genes are spliced, and several genes are expressed as one transcription unit under a single promoter. This insight into THEV's transcriptome may allow the engineering of the VAS to provide immune protection with less or no associated IS. 
\newpage

## INTRODUCTION   
Adenoviruses (AdVs) are non-enveloped icosahedral-shaped DNA viruses, causing infection in virtually all vertebrates. Their double-stranded linear DNA genomes range between 26 and 45kb in size, producing a broad repertoire of transcripts via a highly complex alternative splicing pattern [@Davison2003; @Harrach2008]. The AdV genome is one of the most optimally economized; both the forward and reverse DNA strands harbor protein-coding genes, making it highly gene-dense. There are 16 genes termed "genus-common" that are homologous in all AdVs; these are thought to be inherited from a common ancestor. All other genes are termed "genus-specific". "Genus-specific" genes tend to be located at the termini of the genome while "genus-common" genes are usually central [@Davison2003]. This pattern is observed in *Adenoviridae*, *Poxviridae*, and *Herpesviridae* [@Davison2003; @Upton2003; @McGeoch1999]. The family *Adenoviridae* consists of five genera: *Mastadenovirus* (MAdV), *Aviadenovirus*, *Atadenovirus*, *Ichtadenovirus*, and *Siadenovirus* (SiAdV) [@Harrach2011; @Guimet2016]. Currently, there are three recognized members of the genus SiAdV: frog adenovirus 1, raptor adenovirus 1, and turkey adenovirus 3 also called turkey hemorrhagic enteritis virus (THEV) [@Harrach2011; @Kovács2011; @Davison2000; @Kovács2010; @Katoh2009]. Members of SiAdV have the smallest genome size (~26 kb) and gene content (~23 genes) of all known AdVs, and many “genus-specific” putative genes of unknown functions have been annotated (see __Figure 1__)[@Harrach2008; @Kovács2011; @Davison2003].

@@

Myriads of studies have elucidated the AdV transcriptome in fine detail [@Donovan2020; @Zhao2014]. However, a large preponderance of studies focus on MAdVs - specifically human AdVs - thus, most of the current knowledge regarding AdV gene expression and replication is based on MAdV studies, which is generalized for all other AdVs [@Guimet2016; @Wolfrum2013]. MAdV genes are transcribed in a temporal manner; therefore, genes are categorized into five early transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and one major late unit (MLTU), which generates five families of late mRNAs (L1-L5). An additional gene (UXP or U exon) is located on the reverse strand. The early genes encode non-structural proteins such as enzymes or host cell modulating proteins, primarily involved in DNA replication or providing the necessary intracellular niche for optimal replication while late genes encode structural proteins. The immediate early gene E1A is expressed first, followed by the the delayed early genes, E1B, E2, E3 and E4. Then the intermediate early genes, IVa2 and pIX are expressed followed by the late genes [@Zhao2014; @Donovan2020; @Guimet2016]. MAdV makes an extensive use of alternative RNA splicing to produce a very complex array of mRNAs; all but pIX mRNA undergo at least one splicing event. The MLTU produces over 20 distinct splice variants all of which contain three non-coding exons at the 5'-end (collectively known as the tripartite leader, TPL) [@Donovan2020; @Zhao2014]. There is also an alternate 5' three non-coding exons present in varying amounts on a subset of MLTU mRNAs (known as the x-, y- and z-leaders). Lastly, there is the i-leader exon, which is infrequently included between the second and third TPL exons, and codes for the i-leader protein [@Falvey1983]. Thus, the MLTU produces a complex repertoire of mRNA with diverse 5'-UTRs spliced onto different 3' coding exons which are grouped into five different 3'-end classes (L1-L5). Each transcription unit (TU) contains its own promoter driving the expression of all the array of mRNA transcripts produced via alternative splicing of the genes encoded in the unit[@Zhao2014; @Donovan2020; @Guimet2016]. Almost all AdV mRNAs are generated by the excision of one or more introns and most of these introns are located in the 5' or 3' UTRs of pre-mRNA. Thus the viral introns scarcely interrupt the open reading frames (ORFs) [@Davison2003; @Zhao2014].

@@

## RESULTS   
**Overview of sequencing data and analysis pipeline outputs**     
A previous study by Zeinab *et al* showed that almost all THEV transcripts were detectable beginning at 4 hours [@Aboezz2019]. Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and 72-hours post-infection(h.p.i) to ensure an amply wide time window to sample all transcripts. Our paired-end RNA sequencing (RNA-seq) experiment yielded an average of `r avg_tot_count` million total reads of `r seq_len`bp in length per time-point, which were simultaneously mapped to both the virus (THEV) and host (*M.gallopavo*) genomes using the `Hisat2` [@Pertea2016] alignment program. A total of `r total_mapped` million reads from all time-points mapped to the virus genome; this provided good coverage/depth, leaving no regions unmapped. The mapped reads to the virus genome increased substantially from `r cov_4hr` reads at 4 h.p.i to `r cov_72hr` million reads at 72 h.p.i (__Table 1__, __Figure 2a__). From the mapped reads, we identified an overall total of `r total_unq_jncs_submit %>% nrow() %>% comma(.)` unique THEV splice junctions from all time-points, with splice junctions from the later time-points being supported by significantly more sequence reads than earlier time-points. For example all the `r tot_jun_4` unique junctions at 4 h.p.i had less than 10 reads supporting each one, averaging a mere `r mean_juncreads_4` reads/junction. Conversely, the `r tot_jun_72` unique junctions at 72 h.p.i averaged `r mean_juncreads_72` reads/junction, some junctions having coverage as high as `r max_junread72` reads. The substantial increases in splice junctions and mapping reads to the THEV genome over time denotes an active infection, and correlates with our quantitative PCR (qPCR) assay quantifying the total number of viral genome copies over time (__Figure 2b__). Using `StringTie` [@Pertea2016], an assembler of RNA-seq alignments into potential transcripts, the mapped reads for each time-point were assembled into transcripts using the genomic location of the predicted THEV ORFs as a guide. In the consolidated transcriptome, a composite of all unredundant transcripts from all time points, we counted a total of `r nrow(spliced_gtf)` transcripts all of which are novel, ~~and using 3' Rapid Amplification of cDNA Ends (3'RACE) and other methods, we further identified #### unique splice variants~~. Although some exons in some transcripts match the predicted ORFs exactly, most of our identified exons are longer, spanning multiple predicted ORFs (__Figure 3__). The complete list of unique splice junctions mapped to THEV's genome has been submitted to the National Center for Biotechnology Information Gene Expression Omnibus (http://www.ncbi.nlm.nih.gov/geo) under **accession no. XXXXXX**.

**Changes in THEV splicing profile over time**   
AdV gene expression occurs under exquisite temporal control, supervised by designated promoters for each transcription unit (TU) or region. Each promoter typically produce one or few pre-mRNAs that undergo alternative splicing to yield the manifold repertoire of complex transcripts characteristic of AdVs [@Donovan2020; @Zhao2014]. To evaluate the activity of each promoter over time, Firstly, `Ballgown`, a program for statistical analysis of assembled transcriptomes [@Ballgown] was used to estimate and normalize expression levels of all transcripts for each time point in Fragments Per Kilobase of transcript per Million mapped reads (FPKM) units. Very few unique splice junctions, reads, and transcripts were counted at 4 h.p.i; hence, this time point was excluded in this analysis. 
```{r temporal transcript expression levels}
abn_each_12 <- t_expr_each %>% filter(trxpt_id == "E2: TRXPT_21", timepoint == "12h.p.i") %>% pull(fpkm_trxpt_percent)
```
Individually, TRXPT_21 (DBP) - from the E2 region - was the most significantly expressed at 12 h.p.i, comprising about `r abn_each_12`% of the total transcripts. Transcripts in the E3 and E4 regions also contributed significant proportions, and noticeably, some MLP region transcripts. The later time points were dominated by the MLP region; transcripts TRXPT_10 and TRXPT_14 were most abundantly expressed at 24 and 72 h.p.i, respectively, as expected (__Figure 4a__). When we performed analysis of the FPKM values of transcripts per region we found a similar pattern: the E2 region was the most abundantly expressed at 12 h.p.i, after which the MLP region assumes predominance (__Figure 4b__). Secondly, we estimated relative abundances of all splice junctions for each time point using the raw reads. We counted as significantly expressed only junctions with coverage of at least 1% of the total splice junction reads counted at the given time point. At 12 h.p.i, `r nrow(sig_12_juncs)` junctions meet the 1% threshold, and were comprised of predominantly early region (E1, E2, E3, and E4) junctions, albeit the MLP was the single most preponderant region overall, constituting `r round(most_sig_12, 1)`% of all the junctions reads. (__Table 2a__). The levels of the top most abundant junctions at 12 h.p.i were maintained also 24 h.p.i as the most significantly expressed. However, here, the MLP-derived junctions were unsurprisingly even more preponderant overall, accounting for `r round(most_sig_24, 1)`% of all the junction reads counted (__Table 2b__). At 72 h.p.i, the trend of increased activity of the MLP continued as expected; at this time, the MLP-derived junctions were not only the most abundant overall - accounting for `r round(most_sig_72, 1)`% of all junctions reads - but also contained the most significantly expressed individual junctions (__Table 2c__. Also see __Supplementary Tables 1a-c__; __Figure 4c__). When we limited this analysis to only junctions in the final transcriptome, the relative abundances of the junctions for each region over time was generally similar to the pattern seen with all the junctions included (__Figure 4d__).  
Furthermore, we analyzed splice donor and acceptor site nucleotide usage over time to investigate any peculiarities that THEV may show, generally or over the course of the infection. We found that most splice donor-acceptor sequences were unsurprisingly the canonical GU-AG nucleotides.

**Early Region 1 (E1) transcripts**. This region in MAdVs is the first transcribed after successful entry of the viral DNA into the host cell nucleus, albeit at low levels [@Zhao2014]. The host transcription machinery solely mediates the transcription of this region. After their translation, the E1 proteins in concert with a myriad of host transcription factors activate the other viral promoters [@Guimet2016]. Only two ORFs (ORF1 [sialidase] and Hyd) are predicted in this region; however, we discovered `r as.character(english(nrow(filter(spliced_gtf, region == "E1"))))` novel transcripts in this region, which collectively contain `r nrow(reg_e1_brkdown)` unique splice junctions (__Figure 5__). Most of the encoded proteins of the novel transcripts are distinct from the predicted ORFs, although they all have the potential to encode the Hyd protein as the 3'-most coding sequence (CDS) if the first start codon (SSC) is skipped. The 5'-most CDS of TRXPT_1 is multi-exonic, producing a 17.9 kilodalton (kDa) protein of 160 residues [amino acids (aa)]. The CDS begins in the first exon, starting at position 211, spans the second exon, and terminates in the third exon at position 2312. From the 5'-most SSC, TRXPT_2 encodes the largest protein in this region ——  a 64.3 kDa, 580 aa protein with the same SSC as TRXPT_1 (position 211). This CDS spans almost the entire predicted ORF1 and Hyd, coming short in two regards: it is spliced from 1655 to 1964 (ORF1's C-terminus, including the stop codon), and it's stop codon (STC; position 2312) is 13 bp short of the Hyd STC. However, it has an SSC 102 bp upstream and in-frame with ORF1's predicted SSC. The N-terminus of TRXPT_2 CDS therefore, shares substantial protein sequence similarity with ORF1 but Hyd and TRXPT_2 are not in-frame; hence no protein sequence similarity. TRXPT_3 is almost identical to TRXPT_1, except for the lack of TRXPT_1's second exon. From our analysis, TRXPT_3 and TRXPT_4 seem to have transcription start sites (TSS) downstream of the TSS of TRXPT_1 and TRXPT_2; however, given that studies in MAdVs show that E1 mRNAs share a common 5' and 3' positions, only differing from each other regarding the introns [@Zhao2014], it is likely that TRXPT_3 and TRXPT_4 are incomplete, and the TSS just like the transcription termination site (TTS) are identical for all E1 transcripts. Regardless of the TSS considered for TRXPT_3, the coding potential remains unaffected. Its 5'-most CDS, beginning at 1965 and sharing the same STC as TRXPT_1 and TRXPT_2, produces a 13.1 kDa, 115 residue protein. This CDS (ORF4) was predicted in an earlier study [@Pitcovski1998] but was excluded in later studies [@Beach2009a; @Davison2003]; however, our data suggests it is a bona fide ORF. The coding potential of TRXPT_4 is affected by the TSS considered; if we consider its unmodified TSS, then its coding potential is the same as TRXPT_3 (ORF4 as the first CDS and Hyd as second CDS if the first SSC is skipped). However, if we assume that TRXPT_4 shares the same TSS as TRXPT_1, then the 5'-most CDS is a distinct multi-exonic 15.9 kDa, 143 aa protein with the same SSC as TRXPT_1 and TRXPT_2 but with a unique STC. All splice junctions of the transcripts in this region (except the junction for TRXPT_4) have been validated by cloning and Sanger sequencing of cDNA (__Figure 5b__; __supplementary PCR methods__). Finally, during our validation of TRXPT_2, ORF1 was present on the agarose gel and Sanger sequencing results as a bona fide transcript (__supplementary PCR methods__). This is corroborated by 3'-RACE experiment, which shows transcripts spanning the entire ORF1 and Hyd ORFs without any splicing with a poly-A tail at the TTS of transcripts in this region. The 5'-most CDS of this transcript would encode ORF1. Given that the SSC of the predicted ORF1 is in-frame but downstream of TRXPT_2 SSC, it suggests that the predicted ORF1 CDS is incomplete; it shares the same SSC and TSS, and TTS as TRXPT_2, but has a unique STC.

>>>>

---
title: Elucidating the Transcriptome of Turkey Hemorrhagic Enteritis Virus
output:
  word_document: default
  pdf_document: default
geometry: margin=1in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
bibliography: transcriptome_refs.bib
csl: asm.csl
editor_options:
  chunck_output_type: console
---

@@

\vspace{10mm}

Abraham Quaye^1\*^, Bret Pickett^\*^, Joel S. Griffitts^\*^, Bradford K. Berges^\*^, Brian D. Poole${^\dagger}$^\*^

@@

^\*^Department of Microbiology and Molecular Biology, Brigham Young University\
^1^First-author\
${^\dagger}$ Corresponding Author

**Corresponding Author Information**\
[brian_poole\@byu.edu](mailto:brian_poole@byu.edu)\
Department of Microbiology and Molecular Biology,\
4007 Life Sciences Building (LSB),\
Brigham Young University,\
Provo, Utah\
\newpage

@@

most_sig_72 <- bulk_junc_stats %>% filter(timepoint == "72hpi", region == "MLP") %>% pull(read_count) %>% sum()/as.numeric(tab1$`72h.p.i`[6])*100
```

## ABSTRACT

**Background:** Hemorrhagic enteritis (HE) is a disease affecting 6-12-week-old turkeys characterized by *immunosuppression (IS)* and bloody diarrhea. This disease is caused by *Turkey Hemorrhagic Enteritis Virus (THEV)* of which avirulent strains (THEV-A) that do not cause HE but retain the immunosuppressive ability have been isolated. The THEV-A Virginia Avirulent Strain (VAS) is still used as a live vaccine despite its immunosuppressive properties. *Our objective is to understand the genetic basis by which VAS induces IS*. The *transcriptome of THEV* was studied to set the stage for further experimentation with specific viral genes that may mediate IS.\
**Methods:** After infecting a turkey B-cell line (MDTC-RP19) with the VAS vaccine strain, samples in triplicates were collected at 4-, 12-, 24-, and 72-hours post-infection. Total RNA was subsequently extracted, and poly-A-tailed mRNA sequencing done. After trimming the raw sequencing reads with the `FastQC`, reads were mapped to the THEV genome using `Hisat2` and transcripts assembled with `StringTie`. An in-house script was used to consolidate transcripts from all time-points, generating the final transcriptome. PCR, gel electrophoresis, and Sanger sequencing were used to validate all identified splice junctions.\
**Results and Conclusions:** A total of `r total_mapped` million reads mapped to THEV genome providing good coverage/depth, leaving no regions unmapped. All predicted genes in the genome were represented. In keeping with all adenoviruses, all transcripts were spliced with either with 5'- or 3'-multi exon UTRs hitherto unknown. ~~*Thirteen*~~ novel exons were identified which were validated by PCR and Sanger sequencing. The splicing patterns strongly suggest that there are ~~*three*~~ main promoters (E1, E3, and major late promoters) driving expression of most of the genes with ~~*two*~~ possible minor promoters driving single genes (ORF7 and ORF8). This RNA-sequencing experiment is the first study of THEV gene expression to date. In keeping with other Adenoviruses, almost all THEV genes are spliced, and several genes are expressed as one transcription unit under a single promoter. This insight into THEV's transcriptome may allow the engineering of the VAS to provide immune protection with less or no associated IS. \newpage

## INTRODUCTION

Adenoviruses (AdVs) are non-enveloped icosahedral-shaped DNA viruses, causing infection in virtually all vertebrates. Their double-stranded linear DNA genomes range between 26 and 45kb in size, producing a broad repertoire of transcripts via a highly complex alternative splicing pattern [@Davison2003; @Harrach2008]. The AdV genome is one of the most optimally economized; both the forward and reverse DNA strands harbor protein-coding genes, making it highly gene-dense. There are 16 genes termed "genus-common" that are homologous in all AdVs; these are thought to be inherited from a common ancestor. All other genes are termed "genus-specific". "Genus-specific" genes tend to be located at the termini of the genome while "genus-common" genes are usually central [@Davison2003]. This pattern is observed in *Adenoviridae*, *Poxviridae*, and *Herpesviridae* [@Davison2003; @Upton2003; @McGeoch1999]. The family *Adenoviridae* consists of five genera: *Mastadenovirus* (MAdV), *Aviadenovirus*, *Atadenovirus*, *Ichtadenovirus*, and *Siadenovirus* (SiAdV) [@Harrach2011; @Guimet2016]. Currently, there are three recognized members of the genus SiAdV: frog adenovirus 1, raptor adenovirus 1, and turkey adenovirus 3 also called turkey hemorrhagic enteritis virus (THEV) [@Harrach2011; @Kovács2011; @Davison2000; @Kovács2010; @Katoh2009]. Members of SiAdV have the smallest genome size (\~26 kb) and gene content (\~23 genes) of all known AdVs, and many "genus-specific" putative genes of unknown functions have been annotated (see **Figure 1**)[@Harrach2008; @Kovács2011; @Davison2003].

@@

Myriads of studies have elucidated the AdV transcriptome in fine detail [@Donovan2020; @Zhao2014]. However, a large preponderance of studies focus on MAdVs - specifically human AdVs - thus, most of the current knowledge regarding AdV gene expression and replication is based on MAdV studies, which is generalized for all other AdVs [@Guimet2016; @Wolfrum2013]. MAdV genes are transcribed in a temporal manner; therefore, genes are categorized into five early transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and one major late unit (MLTU), which generates five families of late mRNAs (L1-L5). An additional gene (UXP or U exon) is located on the reverse strand. The early genes encode non-structural proteins such as enzymes or host cell modulating proteins, primarily involved in DNA replication or providing the necessary intracellular niche for optimal replication while late genes encode structural proteins. The immediate early gene E1A is expressed first, followed by the the delayed early genes, E1B, E2, E3 and E4. Then the intermediate early genes, IVa2 and pIX are expressed followed by the late genes [@Zhao2014; @Donovan2020; @Guimet2016]. MAdV makes an extensive use of alternative RNA splicing to produce a very complex array of mRNAs; all but pIX mRNA undergo at least one splicing event. The MLTU produces over 20 distinct splice variants all of which contain three non-coding exons at the 5'-end (collectively known as the tripartite leader, TPL) [@Donovan2020; @Zhao2014]. There is also an alternate 5' three non-coding exons present in varying amounts on a subset of MLTU mRNAs (known as the x-, y- and z-leaders). Lastly, there is the i-leader exon, which is infrequently included between the second and third TPL exons, and codes for the i-leader protein [@Falvey1983]. Thus, the MLTU produces a complex repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced onto different 3' coding exons which are grouped into five different 3'-end classes (L1-L5). Each transcription unit (TU) contains its own promoter driving the expression of all the array of mRNA transcripts produced via alternative splicing of the genes encoded in the unit[@Zhao2014; @Donovan2020; @Guimet2016]. Almost all AdV mRNAs are generated by the excision of one or more introns and most of these introns are located in the 5' or 3' UTRs of pre-mRNA. Thus the viral introns scarcely interrupt the open reading frames (ORFs) [@Davison2003; @Zhao2014].

@@

## RESULTS

**Overview of sequencing data and analysis pipeline outputs**\
A previous study by Zeinab *et al* showed that almost all THEV transcripts were detectable beginning at 4 hours [@Aboezz2019]. Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and 72-hours post-infection(h.p.i) to ensure an amply wide time window to sample all transcripts. Our paired-end RNA sequencing (RNA-seq) experiment yielded an average of `r avg_tot_count` million total reads of `r seq_len`bp in length per time-point, which were simultaneously mapped to both the virus (THEV) and host (*M.gallopavo*) genomes using the `Hisat2` [@Pertea2016] alignment program. A total of `r total_mapped` million reads from all time-points mapped to the virus genome; this provided good coverage/depth, leaving no regions unmapped. The mapped reads to the virus genome increased substantially from `r cov_4hr` reads at 4 h.p.i to `r cov_72hr` million reads at 72 h.p.i (**Table 1**, **Figure 2a**). From the mapped reads, we identified an overall total of `r total_unq_jncs_submit %>% nrow() %>% comma(.)` unique THEV splice junctions from all time-points, with splice junctions from the later time-points being supported by significantly more sequence reads than earlier time-points. For example all the `r tot_jun_4` unique junctions at 4 h.p.i had less than 10 reads supporting each one, averaging a mere `r mean_juncreads_4` reads/junction. Conversely, the `r tot_jun_72` unique junctions at 72 h.p.i averaged `r mean_juncreads_72` reads/junction, some junctions having coverage as high as `r max_junread72` reads. The substantial increases in splice junctions and mapping reads to the THEV genome over time denotes an active infection, and correlates with our quantitative PCR (qPCR) assay quantifying the total number of viral genome copies over time (**Figure 2b**). Using `StringTie` [@Pertea2016], an assembler of RNA-seq alignments into potential transcripts, the mapped reads for each time-point were assembled into transcripts using the genomic location of the predicted THEV ORFs as a guide. In the consolidated transcriptome, a composite of all unredundant transcripts from all time points, we counted a total of `r nrow(spliced_gtf)` transcripts all of which are novel, ~~and using 3' Rapid Amplification of cDNA Ends (3'RACE) and other methods, we further identified \#### unique splice variants~~. Although some exons in some transcripts match the predicted ORFs exactly, most of our identified exons are longer, spanning multiple predicted ORFs (**Figure 3**). The complete list of unique splice junctions mapped to THEV's genome has been submitted to the National Center for Biotechnology Information Gene Expression Omnibus (<http://www.ncbi.nlm.nih.gov/geo>) under **accession no. XXXXXX**.

**Changes in THEV splicing profile over time**\
AdV gene expression occurs under exquisite temporal control, supervised by designated promoters for each transcription unit (TU) or region. Each promoter typically produce one or few pre-mRNAs that undergo alternative splicing to yield the manifold repertoire of complex transcripts characteristic of AdVs [@Donovan2020; @Zhao2014]. To evaluate the activity of each promoter over time, Firstly, `Ballgown`, a program for statistical analysis of assembled transcriptomes [@Ballgown] was used to estimate and normalize expression levels of all transcripts for each time point in Fragments Per Kilobase of transcript per Million mapped reads (FPKM) units. Very few unique splice junctions, reads, and transcripts were counted at 4 h.p.i; hence, this time point was excluded in this analysis.

```{r temporal transcript expression levels}
abn_each_12 <- t_expr_each %>% filter(trxpt_id == "E2: TRXPT_21", timepoint == "12h.p.i") %>% pull(fpkm_trxpt_percent)
```

Individually, TRXPT_21 (DBP) - from the E2 region - was the most significantly expressed at 12 h.p.i, comprising about `r abn_each_12`% of the total transcripts. Transcripts in the E3 and E4 regions also contributed significant proportions, and noticeably, some MLP region transcripts. The later time points were dominated by the MLP region; transcripts TRXPT_10 and TRXPT_14 were most abundantly expressed at 24 and 72 h.p.i, respectively, as expected (**Figure 4a**). When we performed analysis of the FPKM values of transcripts per region we found a similar pattern: the E2 region was the most abundantly expressed at 12 h.p.i, after which the MLP region assumes predominance (**Figure 4b**). Secondly, we estimated relative abundances of all splice junctions for each time point using the raw reads. We counted as significantly expressed only junctions with coverage of at least 1% of the total splice junction reads counted at the given time point. At 12 h.p.i, `r nrow(sig_12_juncs)` junctions meet the 1% threshold, and were comprised of predominantly early region (E1, E2, E3, and E4) junctions, albeit the MLP was the single most preponderant region overall, constituting `r round(most_sig_12, 1)`% of all the junctions reads. (**Table 2a**). The levels of the top most abundant junctions at 12 h.p.i were maintained also 24 h.p.i as the most significantly expressed. However, here, the MLP-derived junctions were unsurprisingly even more preponderant overall, accounting for `r round(most_sig_24, 1)`% of all the junction reads counted (**Table 2b**). At 72 h.p.i, the trend of increased activity of the MLP continued as expected; at this time, the MLP-derived junctions were not only the most abundant overall - accounting for `r round(most_sig_72, 1)`% of all junctions reads - but also contained the most significantly expressed individual junctions (**Table 2c**. Also see **Supplementary Tables 1a-c**; **Figure 4c**). When we limited this analysis to only junctions in the final transcriptome, the relative abundances of the junctions for each region over time was generally similar to the pattern seen with all the junctions included (**Figure 4d**).\
Furthermore, we analyzed splice donor and acceptor site nucleotide usage over time to investigate any peculiarities that THEV may show, generally or over the course of the infection. We found that most splice donor-acceptor sequences were unsurprisingly the canonical GU-AG nucleotides.

**Early Region 1 (E1) transcripts**. This region in MAdVs is the first transcribed after successful entry of the viral DNA into the host cell nucleus, albeit at low levels [@Zhao2014]. The host transcription machinery solely mediates the transcription of this region. After their translation, the E1 proteins in concert with a myriad of host transcription factors activate the other viral promoters [@Guimet2016]. Only two ORFs (ORF1 [sialidase] and Hyd) are predicted in this region; however, we discovered `r as.character(english(nrow(filter(spliced_gtf, region == "E1"))))` novel transcripts in this region, which collectively contain `r nrow(reg_e1_brkdown)` unique splice junctions (**Figure 5**). Most of the encoded proteins of the novel transcripts are distinct from the predicted ORFs, although they all have the potential to encode the Hyd protein as the 3'-most coding sequence (CDS) if secondary start codon usage is considered. The 5'-most CDS of TRXPT_1 is multi-exonic, producing a 17.9 kilodalton (kDa) protein of 160 residues [amino acids (aa)]. The CDS begins in the first exon, starting at position 211, spans the second exon, and terminates in the third exon at position 2312. From the 5'-most start codon (SSC), TRXPT_2 encodes the largest protein in this region - a 64.3 kDa, 580 aa protein with the same SSC as TRXPT_1 (position 211). This CDS spans almost the entire predicted ORF1 and Hyd, coming short in two regards: it is spliced from 1655 to 1964 (ORF1's C-terminus, including the stop codon), and it's stop codon (STC; position 2312) is 13 bp short of the Hyd STC. However, it has an SSC 102 bp upstream and in-frame with ORF1's predicted SSC. The N-terminus of TRXPT_2 CDS therefore, shares substantial protein sequence similarity with ORF1 but Hyd and TRXPT_2 are not in-frame; hence no protein sequence similarity. TRXPT_3 is almost identical to TRXPT_1, except for the lack of TRXPT_1's second exon. From our RNA-seq data, TRXPT_3 and TRXPT_4 seem to have transcription start sites (TSS) downstream of the TSS of TRXPT_1 and TRXPT_2; however, given that studies in MAdVs show that E1 mRNAs share a common 5' and 3' positions, only differing from each other regarding the introns [@Zhao2014], it is likely that TRXPT_3 and TRXPT_4 are incomplete, and the TSS (at position 54bp) just like the transcription termination site (TTS; at position 2325bp) are identical for all E1 transcripts. Regardless of the TSS considered for TRXPT_3, the coding potential remains unaffected. Its 5'-most CDS, beginning at 1965 and sharing the same STC as TRXPT_1 and TRXPT_2, produces a 13.1 kDa, 115 residue protein. This CDS (ORF4) was predicted in an earlier study [@Pitcovski1998] but was excluded in later studies [@Beach2009a; @Davison2003]; however, our data suggests it is a bona fide ORF. The coding potential of TRXPT_4 is affected by the TSS considered; if we consider its unmodified TSS, then its coding potential is the same as TRXPT_3 (ORF4 as the first CDS and Hyd as second CDS if the first SSC is skipped). However, if we assume that TRXPT_4 shares the same TSS as TRXPT_1, then the 5'-most CDS is a distinct multi-exonic 15.9 kDa, 143 aa protein with the same SSC as TRXPT_1 and TRXPT_2 but with a unique STC. All splice junctions of the transcripts in this region (except the junction for TRXPT_4) have been validated by cloning and Sanger sequencing of cDNA (**Figure 5b**; **supplementary PCR methods**). Finally, during our validation of TRXPT_2, ORF1 was present on the agarose gel and Sanger sequencing results as a bona fide transcript (**supplementary PCR methods**). This is corroborated by 3'-RACE experiment, which shows transcripts spanning the entire ORF1 and Hyd ORFs without any splicing with a poly-A tail immediately after the TTS of transcripts in this region. The 5'-most CDS of this transcript would encode ORF1. Given that the SSC of the predicted ORF1 is in-frame but downstream of TRXPT_2 SSC, it suggests that the predicted ORF1 CDS is incomplete; it shares the same SSC and TSS, and TTS as TRXPT_2, but has a unique STC.
